
layout: topic
title: "Data exploration"
output: html_document
---

**Assigned Reading:**

> Zuur, A. F., E. N. Ieno, and C. S. Elphick. 2010. A protocol for data exploration to avoid common statistical problems. *Methods in Ecology and Evolution* **1**: 3-14.  [DOI: 10.1111/j.2041-210X.2009.00001.x](https://dx.doi.org/10.1111/j.2041-210X.2009.00001.x)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# # Define the directory where images generated by knit will be saved
# knitr::opts_chunk$set(fig.path = "images/03-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
# repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## Key takeaways from Zuur et al. (2010):

Things we should explore and check per Zuur et al. (2010). Note: we don't have to do every step every time!


1. Outliers Y & X
    + Outliers in response variable vs in covariates - to be dealt with differently
    + Transformation less desirable for response variable than in covariates?
2. Homogeneity Y
3. Normality Y
    + Example of importance of biological intuition and graphical investigation
        + Transformation to get normality may not be desirable--will learn more when we discuss generalized linear models.
4. Zero trouble Y
    + Zero-inflated GLM
    + Double zeros, or joint absences - what do they mean? e.g., spatial clumping
    + We will not be covering multivariate analyses in this course! (but they're super cool, so go check them out)
5. Collinearity X
    + VIF of 10, 3, or 2 - reason for these values?
    + VIFs, or common sense or biological knowledge
    + Also... are pairwise comparisons worthwhile?
6. Relationships Y & X
    + Multi-panel scatter plots for checking for outliers
7. Interactions
    + Are data balanced?
    + coplots (i.e., ggplot2::facet_wrap)
8. Independence Y
    + ACF, variograms, Moran's I for checking for temporal and spatial non-independence
    + We will focus on this in our Autocorrelation Topic.
    

### BUT: Biological intuition = key to make decisions about stats

+ Hypothesis testing vs hypothesis generation
    + One solution - two data sets - one to create hypotheses and one to test them
    + But only practical for large data sets
+ Graphical tools very powerful, but other summaries of the data are also very powerful:
    + What is the 'central tendency?'
    + How many replicates do you have, etc.?


## Data exploration examples:

The code below is built on data from Roberts et al. (2022) in *Ecological Solutions and Evidence*.

```{r, include = TRUE, eval = TRUE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse, cowplot, quiet = TRUE)

# Load Roberts et al. (2022) data:
dat_grass <- 
  read.csv("https://raw.githubusercontent.com/LivingLandscapes/LargeScaleFireRestoresGrasslandBirdRichness/LargeScaleFireRestoresGrasslandBirdRichness/LargeScaleFireRestoresGrasslandBirdRichness_RProj/LoessCanyons_BBS_Data/LoessCanyonsBBS_DataRaw.csv")

##### Data column descriptions: 
# - Route: name of breeding bird survey route. 
# - Stop: ID for each stop along survey route+
# - Year: year of survey
# - Rich_Grass: grassland bird species richness 
# - Route_factor: name of breeding bird survey route. 
# - Year_Num: survey years re-numbered from 1 (2010) to 7 (2016)
# - Burned: if this stop was burned on or after the given year, 1, else 0
# - Year_Burned: the year that a given stop was burned
# - easting/northing: easting/northing coordinates in UTM 
# - count: number of 30x30m pixels within 400m of each stop. Number of pixels
#          vary because some are masked because they were not 'rangeland' 
#          pixels. See Methods/Data Collection/Tree Cover for details.
# - mean: mean percent tree cover across all 30x30m pixels within 400m of
#         each stop
# - stdDev: standard deviation of percent tree cover across all 30x30m pixels 
#           within 400m of each stop
# - TSF: years-since-fire

```

### Checking for outliers

Zuur et al. recommends plotting your data using boxplots and dotcharts to detect outliers. Violin plots are also great options. Before dealing with suspected outliers, make sure they are actually outliers!

*Important*: If you have outliers, removing them is likely not the first thing you should do. There may be other error distribution options you could use or transformations. It also could be a hint that you haven't gotten a sufficiently large sample size to flesh out your population's distribution

**Boxplot**

```{r, eval = TRUE, render= TRUE, echo = FALSE}

ggplot(data = dat_grass,
       mapping = aes(y = Rich_Grass, 
                     x = as.factor(Year))) + 
  geom_boxplot() +
  ylab("Grassland Bird Richness") + 
  xlab("Year")

```

**Violin plots**

Violin plots show more data distribution details, but they can be messy. These are conditional (by year) and display the 10th, 50th (i.e., median), and 90th quantiles as horizontal lines.

```{r, eval = TRUE, render= TRUE, echo = FALSE}

ggplot(data = dat_grass,
       mapping = aes(x = as.factor(Year), y = Rich_Grass, group = Year)) + 
  geom_violin(draw_quantiles = c(0.1, 0.5, 0.9)) +
  ylab("Grassland Bird Richness") + 
  xlab("Year")

```

**Dotchart for multiple variables**

I personally don't use these as much, but they can be useful.

```{r, eval = TRUE, include = TRUE}

dat_grass %>%
  arrange(Rich_Grass) %>%
  mutate(rowID = 1:n()) %>%
  ggplot() +
    facet_wrap(~ Year) +
    geom_point(mapping = aes(x = Rich_Grass, y = rowID, group = Year)) +
    xlab("Grassland Bird Richness")

```



### What's your data's distribution?

Some statistical tests assume normal distributions, making it important to check the shape of your data. We will learn about probability distributions in our Probability Distributions Topic.

```{r, eval = TRUE, include = TRUE}

ggplot() +
    geom_histogram(data = dat_grass,
                   mapping = aes(x = Rich_Grass),
                   binwidth = 1) +
  ylab("Frequency") +
  xlab("Grassland Bird Richness")

```

### Collinearity?

It's absolutely important to check for pairwise correlations, which we can do with the "cor()" function as below. I typically use the default "Pearson" method, but you should read the R Documentation for other options (kendall, spearman). However, pairwise correlations should be taken with a grain of salt. See this quote from the R Documentation for the "performance::check_collinearity()" function:

"*Multicollinearity should not be confused with a raw strong correlation between predictors... Remember: "Pairwise correlations are not the problem. It is the conditional associations - not correlations - that matter." (McElreath 2020, p. 169)*"

We will talk about Zuur's preference of "variance inflation factors" to check for collinearity in the linear models review Topic.

```{r, eval = TRUE, include = TRUE}

cor(dat_grass[ , c("Year", "Burned", "mean", "stdDev", "TSF")])

```


### Is there actually a relationship between X and Y?!

For this, we want to plot the response/dependent variable (grassland bird richness) against potential predictor/independent variables. Here, I use quick-and-dirty generalized additive models per ggplot2::geom_smooth and then combine the plots with cowplot::plot_grid

```{r, eval = TRUE, include = TRUE}

# Making individual plots
rich_treeMean <-
  dat_grass %>%
  ggplot(aes(x = mean, y = Rich_Grass)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x)) + # a simple generalized additive model!
  ylab("Grassland Bird Richness") +
  xlab("Mean % Tree Cover") +
  theme_classic() # FYI: there are lots of fun pre-made themes in ggplot2
rich_TSF <-
  dat_grass %>%
  ggplot(aes(x = TSF, y = Rich_Grass)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x)) +
  ylab("Grassland Bird Richness") +
  xlab("Years-since-fire") +
  theme_bw() # Another theme

# Combine plots
cowplot::plot_grid(rich_treeMean, rich_TSF, ncol = 2)

```

Also, we can use the "pairs()" function to create a bunch of scatterplots:

```{r, eval = TRUE, include = TRUE}

pairs(dat_grass[ , c("Year", "Burned", "mean", "stdDev", "TSF")],
      lower.panel = NULL)

```

### Should we consider interactions between predictor variables?

Use ggplot2::facet_wrap alongside ggplot2::stat_smooth to check for potential interactions between mean tree cover and Year:

```{r, eval = TRUE, include = TRUE}

# Making individual plots
dat_grass %>%
  ggplot(aes(x = mean, y = Rich_Grass)) +
  facet_wrap(~ Year) + # Make facet plots by Years-since-fire
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x) +
  ylab("Grassland Bird Richness") +
  xlab("Mean % Tree Cover") +
  theme_minimal()

```

---


## Lab questions and exercises:


**Q1:** When should you let go of an outlier? Are there outliers in the example data?


**Q2:** What are the minimum and maximum values (i.e., the 'range') for each predictor variable? What do you notice about the ranges?


**Q3:** Are there other potential interactions we should consider in the example data? Recreate some ggplot2::facet_wrap plots to see for yourself.


**Q4:** When is it appropriate to transform the response variable?

<!-- > "There are three main reasons for a transformation: **to reduce the effect of outliers** (especially in covariates), **to stabilize the variance** and **to linearize relationships**. However, using more advanced techniques like GLS and GAMs, heterogenity and nonlinearity problems can be solved, making transformation less important." -->


**Q5:** Create some more plots or data summaries exploring the response variable, grassland bird richness (Rich_Grass). What do you notice about the distribution, the spread (i.e., standard deviation and variance), the central tendencies, etc.? Does this variable fit the "normality" assumption?


**Q6:** Are the data 'balanced' in terms of sampling? Create figure(s) or table(s) to answer this.