---
layout: topic
title: "Data exploration"
output: html_document
---

**Assigned Reading:**

> Zuur, A. F., E. N. Ieno, and C. S. Elphick. 2010. A protocol for data exploration to avoid common statistical problems. *Methods in Ecology and Evolution* **1**: 3-14.  [DOI: 10.1111/j.2041-210X.2009.00001.x](https://dx.doi.org/10.1111/j.2041-210X.2009.00001.x)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# # Define the directory where images generated by knit will be saved
# knitr::opts_chunk$set(fig.path = "images/03-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
# repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## Key takeaways from Zuur et al. (2010):

Things we should explore and check per Zuur et al. (2010). Note: we don't have to do every step every time!


1. Outliers Y & X
    + Outliers in response variable vs in covariates - to be dealt with differently
    + Transformation less desirable for response variable than in covariates?
2. Homogeneity Y
3. Normality Y
    + Example of importance of biological intuition and graphical investigation
        + Transformation to get normality may not be desirable--will learn more when we discuss generalized linear models.
4. Zero trouble Y
    + Possibility of complete separation--which is problematic for binary response variables.
    + Can (sometimes) be dealt with by using a zero-inflated error distribution.
    + We will not be covering multivariate analyses in this course! (but they're super cool, so go check them out)
5. Collinearity X
    + VIFs, or common sense or biological knowledge
    + Also... are pairwise comparisons worthwhile?
6. Relationships Y & X
    + Multi-panel scatter plots useful here.
    + But nonlinearities can obfuscate relationships.
7. Interactions
    + Are data balanced?
    + coplots (i.e., ggplot2::facet_wrap)
8. Independence Y
    + ACF, variograms, Moran's I for checking for temporal and spatial non-independence
    + We will focus on this in our Autocorrelation Topic.
    

### BUT: Biological intuition = key to make decisions about stats

+ Hypothesis testing vs hypothesis generation
    + One solution - two data sets - one to create hypotheses and one to test them
    + But only practical for large data sets
+ Graphical tools very powerful, but other summaries of the data are also very powerful:
    + What is the 'central tendency?'
    + How many replicates do you have, etc.?


## Data exploration examples:

The code below is built on data from Roberts et al. (2022) in *Ecological Solutions and Evidence*.

```{r, include = TRUE, eval = TRUE, warning=FALSE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse, cowplot, sf, quiet = TRUE)

# Load Roberts et al. (2022) data:
dat_grass <- 
  read.csv("https://raw.githubusercontent.com/LivingLandscapes/LargeScaleFireRestoresGrasslandBirdRichness/LargeScaleFireRestoresGrasslandBirdRichness/LargeScaleFireRestoresGrasslandBirdRichness_RProj/LoessCanyons_BBS_Data/LoessCanyonsBBS_DataRaw.csv")

##### Data column descriptions: 
# - Route: name of breeding bird survey route. 
# - Stop: ID for each stop along survey route+
# - Year: year of survey
# - Rich_Grass: grassland bird species richness 
# - Route_factor: name of breeding bird survey route. 
# - Year_Num: survey years re-numbered from 1 (2010) to 7 (2016)
# - Burned: if this stop was burned on or after the given year, 1, else 0
# - Year_Burned: the year that a given stop was burned
# - easting/northing: easting/northing coordinates in UTM 
# - count: number of 30x30m pixels within 400m of each stop. Number of pixels
#          vary because some are masked because they were not 'rangeland' 
#          pixels. See Methods/Data Collection/Tree Cover for details.
# - mean: mean percent tree cover across all 30x30m pixels within 400m of
#         each stop
# - stdDev: standard deviation of percent tree cover across all 30x30m pixels 
#           within 400m of each stop
# - TSF: years-since-fire

```

### Outliers, zero trouble, balance

Zuur et al. recommends plotting your data using boxplots and dotcharts to detect outliers. Violin plots and histograms are also great options. I don't personally use dotcharts... because I see the point of them.

*Important*: If you have outliers, removing them is likely not the first thing you should do. There may be other error distribution options you could use or transformations. It also could be a hint that you haven't gotten a sufficiently large sample size to flesh out your population's distribution

#### Boxplot

```{r, eval = TRUE, render= TRUE, echo = FALSE}

ggplot(data = dat_grass,
       mapping = aes(y = Rich_Grass)) + 
  geom_boxplot() +
  ylab("Grassland Bird Richness") + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())

```

#### Violin plot

Violin plots show more data distribution details, but they can be busy and tough to interpret. These are conditional (by year) and display the 10th, 50th (i.e., median), and 90th quantiles as horizontal lines.

```{r, eval = TRUE, render= TRUE, echo = FALSE}

ggplot(data = dat_grass,
       mapping = aes(x = as.factor(Year), y = Rich_Grass, group = Year)) + 
  geom_violin(draw_quantiles = c(0.1, 0.5, 0.9)) +
  ylab("Grassland Bird Richness") + 
  xlab("Year")

```

#### Histogram

Histograms show frequency of values and the shape or distribution of the values. Here's a histogram of our response variable:

```{r, eval = TRUE, include = TRUE}

ggplot() +
    geom_histogram(data = dat_grass,
                   mapping = aes(x = Rich_Grass),
  # Notice we're specifying the binwidth. Why should (or shouldn't) we do this?
                   binwidth = 1) + 
  ylab("Frequency") +
  xlab("Grassland Bird Richness") + 
  # Making it prettier 
  theme_bw() + 
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

```

> **Challenge 1:** 
>
> 1. Check the minimum and maximum values (i.e., the 'range') for the response variable and each predictor variable. What do you notice about the ranges? Are we going to run into "zero trouble" with the response variable?
>
> 2. Create histograms for i) mean tree cover, ii) TSF, iii) and standard deviation in tree cover. **BONUS:** Make them all in a single ggplot call.
>
> 3. Check for outliers in the response variable (grassland bird species richness) and predictor variables.
>
> 4. Are the data 'balanced' in terms of sampling? Create figure(s) or table(s) to answer this. Some of the figures you've already created may also help...

### Normality and Homogeneity in Y 

Some statistical tests assume your response variable conforms to a normal distribution. We will learn about probability/error distributions in our Probability Distributions Topic. But for now, let's forge ahead and make a Q-Q plot (quantile-quantile plot):

```{r, eval = TRUE, include = TRUE}

ggplot(data = dat_grass,
       mapping = aes(sample = Rich_Grass)) +
  geom_qq() +
  geom_qq_line(color = "red") + 
  ylab("Grassland Bird Richness") +
  xlab("Theoretical Quantiles")

```

> **Challenge 2:** 
>
> Does the response variable fit the "normality" assumption?
>

### Collinearity, relationships between X and Y

It's absolutely important to check for pairwise correlations, which we can do quickly with the `pairs` and `cor` functions. For the `cor` function, I typically use the default "Pearson" method, but you should read the R Documentation for other options (kendall, spearman). However, pairwise correlations should be taken with a grain of salt. See this quote from the R Documentation for the "performance::check_collinearity()" function:

"*Multicollinearity should not be confused with a raw strong correlation between predictors... Remember: "Pairwise correlations are not the problem. It is the conditional associations - not correlations - that matter." (McElreath 2020, p. 169)*"

We will talk about Zuur's preference of "variance inflation factors" to check for collinearity in the linear models review Topic.

> **Challenge 3:** 
>
> 1. Check for pairwise correlations between predictor variables graphically using the `pairs` function.
> 
> 2. Check for pairwise correlations between predictor variables statistically using the `cor` function.
> 
> 3. Are there "linear" relationships between the response variable and predictor variables? How can you use output from the `pairs` function to tell?


### Is there actually a relationship between X and Y?!

Let's also plot the response/dependent variable (grassland bird richness) against potential predictor/independent variables. Here, I use quick-and-dirty generalized additive models per ggplot2::geom_smooth and then combine the plots with cowplot::plot_grid

```{r, eval = TRUE, include = TRUE}

# Making individual plots
rich_treeMean <-
  dat_grass %>%
  ggplot(aes(x = mean, y = Rich_Grass)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x)) + # a simple generalized additive model!
  ylab("Grassland Bird Richness") +
  xlab("Mean % Tree Cover") +
  theme_classic() # FYI: there are lots of fun pre-made themes in ggplot2
rich_TSF <-
  dat_grass %>%
  ggplot(aes(x = TSF, y = Rich_Grass)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x)) +
  ylab("Grassland Bird Richness") +
  xlab("Years-since-fire") +
  theme_bw() # Another theme

# Combine plots
cowplot::plot_grid(rich_treeMean, rich_TSF, ncol = 2)

```


### Interactions between predictor variables

Two-way interactions are what we're looking for here. There is (technically) no limit to how many interactions you can have, but I find it gets super confusing to interpret interactions past three-way interactions--and even those are crazy for me.

> **Challenge 4:** 
>
> 1. Are there interactions between predictor variables that we should consider? Recreate some ggplot2::facet_wrap plots to check.


### Bonus Round: Spatial Patterns

> **Bonus Challenge** 
>
> Map grassland bird species richness by year. Do you see any spatial patterns? Use the `ggplot` and `sf` packages to do this!

``` {r, include = TRUE, eval = TRUE}

# To get you started, run this code to transform our data.frame into an `sf` object:
dat_sf <- 
  st_as_sf(dat_grass,
           coords = c("easting", "northing"),
           crs = st_crs(32614)) # This is the code for UTM Zone 14

```

